---
title: "Morphoditar package use case: Analysis of lyrics"
author: "Michael Škvrňák <michael.skvrnak@rozhlas.cz>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette demonstrates functions contained in the *morphoditar* R package which 
accesses the [MorphoDiTa](http://lindat.mff.cuni.cz/services/morphodita/run.php) 
(Morphological Dictionary and Tagger) API developed by Institute of Formal and Applied
Linguistics, Faculty of Mathematics and Physics, Charles University in Prague.

The demonstration is performed using lyrics of a non-random selection of Czech bands 
from the local do-it-yourself scene with genres ranging from hardcore-punk to indie pop[^1]. 
In particular, it includes the bands:  

* [Dukla](http://dukla.bandcamp.com/),  
* [Esazlesa](https://esazlesa.bandcamp.com/),  
* [Gattaca](https://gattaca.bandcamp.com/),  
* [±0](https://plusminusnula.bandcamp.com/),  
* [Remek](https://remekdiy.bandcamp.com/), and  
* [Role](https://rolerolerole.bandcamp.com/)

Please note that MorphoDiTa is available for non-commercial purposes only.

## Motivation

Research on psychology of language indicate that the choice of words is associated with
psychological state of authors. For instance, according to [Tausczik and Pennebaker (2010)](http://journals.sagepub.com/doi/abs/10.1177/0261927X09351676?journalCode=jlsa): 
*People who are experiencing physical or emotional pain tend to have their attention drawn* 
*to themselves and subsequently use more first-person singular pronouns*. 

Also, *"Pronouns and verb tense are useful linguistic elements that can help identify focus,*
*which, in turn, can show priorities, intentions, and processing. Some care should be *
*taken in evaluating how pronouns and verbs are used. An exception to the pronoun-*
*attention rule concerns first-person plural pronouns—“we,” “us,” and “our.” Sometimes *
*“we” can signal a sense of group identity, such as when couples are asked to evaluate *
*their marriages to an interviewer, the more the participants use “we,” the better their *
*marriage"*

Thus, it does not seem to be an overstretch to assume that music bands playing different
genres will use pronouns in different frequencies as their lyrics should focus on different
priorities. 

To quote, a hardcore punk band not included in the analysis [Antisocial Skills](https://antisocialskills.bandcamp.com/releases), 
in particular their song Different frequencies:  
  
*We just we have different priorities*  
*We work on different frequencies*  

In general, we could assume that the attention which bands playing music of different genres
differs in respect to what they perceive as the thing that sucks. To put it bluntly, 
indie pop bands tend to sing songs like *"you don't love me and it sucks"*, 
in contrast to hardcore punk bands which tend to focus on flaws of 
*"the society/consumerism/capitalism which sucks"*. 

Based on this, we can formulate tentative hypotheses like:  

* Indie pop/something bands use first-person and second-person singular pronouns more often than hardcore bands.
* Hardcore bands use first-person and third-person plural pronouns more often than indie pop/something bands.


## Analysis

(If you don't care about the programming stuff, scroll down to colourful charts).

### Processing the data
First install the necessary packages and pull docker image in which run the
MorphoDiTa. 
Let's start with loading the data. They are already stored in the package, 
so we need to call the `data` function.
```{r}
library(devtools)
devtools::install_github("skvrnami/mordor")
library(mordor)
system("docker pull skvrnami/mordor-docker:latest")
system("docker run -d -p 4000:80 skvrnami/mordor-docker")
library(dplyr)
library(ggplot2)

data("dukla")
data("esazlesa")
data("gattaca")
data("plusminusnula")
data("remek")
data("role")

str(dukla)
```

### morphoditaR functions

Before continuing, let's clarify what functions are defined in the package. 

During processing of the data, the first function needed is the `tag_morphodita` 
function which send requests to the API and returns data (tagged words). 
More specifically, it returns S3 object with attributes specifying the request 
(URL, model, tagset, and language) and data.frame with the output from the API.

```{r}
out <- tag_morphodita("To je život")
str(out)
out
```

Clearly, the API does not return tags in a form which would enable to work with them
easily. Therefore, another function splits the tags into separate columns.  
In the case of Czech language models, the tags can be separated into categories specified
[in this document](https://ufal.mff.cuni.cz/pdt/Morphology_and_Tagging/Doc/hmptagqr.html), 
and in the case of English language model, the tags are specified [here](https://ling.upenn.edu/courses/Fall_2003/ling001/penn_treebank_pos.html).

```{r}
out1 <- out %>% split_tags
out1$output
```

The categories, however, are not very comprehensible and consulting the categories
with the documents mentioned above is not user-friendly. Thus, there is `recode_tags`
function which gives the tags proper labels.

```{r}
out2 <- out1 %>% recode_tags
out2$output
```

We can continue in the analysis by defining helper functions which tag the lyrics 
and split the returned tags into columns and recode them using the functions mentioned above.
Also, we need to define function that will create a dataset containing the tagged lyrics
together with data on the band, album, and song the lyrics come from.

```{r}
# Define helper functions

## Tag lyrics
tag_data <- function(df){
    lapply(df$lyrics, function(x) x %>% tag_morphodita %>% split_tags %>% recode_tags)
}

## Add metadata to the tagged output
add_metadata <- function(output, source_lyrics, band){
    for (i in seq_len(length(output))){
        output[[i]]$output$song <- source_lyrics$song[i]
        output[[i]]$output$album <- source_lyrics$album[i]
        output[[i]]$output$band <- band
    }
    
    do.call(rbind, lapply(output, function(x) x[["output"]]))
}

```

Then, we process all of the datasets with lyrics and merge them together into single
dataset.
```{r}

dukla_lyrics <- add_metadata(tag_data(dukla), dukla, "Dukla")
esazlesa_lyrics <- add_metadata(tag_data(esazlesa), esazlesa, "Esazlesa")
gattaca_lyrics <- add_metadata(tag_data(gattaca), gattaca, "Gattaca")
plusminusnula_lyrics <- add_metadata(tag_data(plusminusnula), plusminusnula, "±0")
remek_lyrics <- add_metadata(tag_data(remek), remek, "Remek")
role_lyrics <- add_metadata(tag_data(role), role, "Role")

all_lyrics <- do.call(rbind, lapply(apropos("_lyrics"), get))
all_lyrics$genre <- ifelse(all_lyrics$band %in% c("Esazlesa", "Gattaca", "Remek"), 
                           "hardcore", "indie something")
```

The resulting dataset looks like this:
```{r}
str(all_lyrics)
```

#### Personal pronouns
Then we can focus on personal pronouns (já, ty, on/ona/ono, my, vy, oni/ony) contained
in the lyrics.
```{r}
personal_pronouns <- subset(all_lyrics, SUBPOS == "Personal pronoun, clitical (short) form" | 
                                        SUBPOS == "Personal pronoun")
personal_pronouns$type <- paste(personal_pronouns$NUMBER, personal_pronouns$PERSON)

pronouns_sum <- personal_pronouns %>% group_by(type, band, genre) %>% summarise(length = n())
```

```{r fig.width=7, fig.height=5}
ggplot(pronouns_sum, aes(x = band, y = length, fill = type)) + 
    geom_bar(stat = "identity") + facet_grid(. ~ genre, scales = "free") + 
    labs(title = "Personal pronouns", y = "frequency of pronouns")
```

#### Possessive pronouns
```{r}
possessive_pronouns <- subset(all_lyrics, 
                              SUBPOS == "Possessive pronoun 'můj', 'tvůj', 'jeho/její'")
possessive_pronouns$type <- paste(possessive_pronouns$POSNUMBER, possessive_pronouns$PERSON)

possessive_sum <- possessive_pronouns %>% group_by(type, band, genre) %>% summarise(length = n())

```

```{r fig.width=7, fig.height=5}
ggplot(possessive_sum, aes(x = band, y = length, fill = type)) + 
    geom_bar(stat = "identity") + facet_grid(. ~ genre, scales = "free") + 
    labs(title = "Possessive pronouns", y = "frequency of pronouns")
```

## Interpretation

So the distinction between the genres is not that clear-cut as one could expect. 
The most common personal pronoun is first-person singular for all bands except for Esazlesa
which use second-person singular more often. Their use of second-person singular also 
present a deviation from the hypothesis. 

We can explain this deviation by the fact that the underlying mechanism which should drive
the selection of personal pronouns does not always hold. Hardcore bands are not predestined
to write about despair caused by modern societies, but they can also sing about failures within
inter-personal relationships as they do in [For Better or Worse/V dobrém i ve zlém](https://esazlesa.bandcamp.com/track/v-dobr-m-i-zl-m) 
which is about failed marriage.  
In addition, some of their lyrics are not available on Bandcamp, such as the lyrics of 
[Middle Children of History/Průměrný děti historie](https://esazlesa.bandcamp.com/track/pr-m-rn-d-ti-historie) where they quote 
Palahniuk's Fight Club/Tyler Durden:  
*"Advertising has __us__ chasing cars and clothes, working jobs __we__ hate so __we__ can buy shit __we__ don’t need. __We’re__ the middle children of history, man. No purpose or place. __We__ have no Great War. No Great Depression. __Our__ Great War’s a spiritual war… __our__ Great Depression is __our__ lives."*  
(Although, the English translation contain much more personal pronouns than the Czech translation.
Therefore, exploring the conjugation of verbs might be another possible approach how to 
investigate the difference between genres/authors.)

Apart from that, the use of plural pronouns merely supports the hypothesis. 

If you miss statistical tests, you can use *"perverse language of statistics and numbers"* ([Gattaca: Workfare](https://gattaca.bandcamp.com/track/workfare-workfare-2)) to compute 
them yourself. DIY, d'oh. 

[^1]: More specifically, the selection was based on the author's perception that the bands don't suck and they have lyrics on bandcamp so that it could be scraped. 
